---
title: "reverse_engineering_02_Teo"
author: "Teona"
date: "2025-03-10"
output: html_document
---

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(openxlsx)
library(readxl)

```


```{r}
payouts_data <- clean_names(read_excel("input/payouts.xlsx"))

print(payouts_data)
```


"The Times analyzed a city database listing more than 18,000 payouts dating back to July 2005, obtained through a public records request." The problem is that there are no official numbers from before 2005, so reporters can’t really check if what officials are saying is true or not.


#Q1: What are the total legal payout costs in L.A. for each year, and which year had the highest payout?
```{r}
expenses_by_year <- payouts_data %>%
  group_by(fiscal_year) %>%
  summarize(amount = n()) %>%
  arrange(desc(amount))


print(expenses_by_year)
```

Article is published in 2018. 

"Last budget year (2017), the city paid out more than $200 million in legal settlements and court judgments — a record amount that was more than the city spent on its libraries or fixing its streets."

#Q2: 

- How much did the city spend on legal settlements and court judgments in 2017?
- What was the largest amount the city has ever paid out in legal settlements and judgments in a single year?
- How did the city's spending on legal settlements and court judgments surpass its spending on libraries or fixing its streets?

```{r}
#2.1

legal_settlement_payments <- payouts_data %>%
  filter(fiscal_year == 2017) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE))

print(legal_settlement_payments)

#2.2

largest_settlement_payment <- expenses_by_year %>%
   arrange(desc(amount))


print(largest_settlement_payment)


#2.3

lib_str_services_spending <- payouts_data %>%
  filter(department == 'Library' | department == 'Public Works / Bureau of Street Services') %>%
  summarize(total_amount = sum(amount, na.rm = TRUE))

print(lib_str_services_spending)
  
  

#compare spending???

```



#Q3: 



```{r}

```


"Thousands of legal battles involving the Los Angeles Police Department, including a rash of recent settlements over police shootings and wrongful imprisonment, have added up to more than 40% of the total — a higher sum than any other city department. That includes more than $24 million paid to two men who were freed after decades in prison, along with lawsuits over arrests, injuries, car crashes and alleged retaliation and discrimination against its own employees."

#Q4: 

- What percentage of the city’s total legal payouts are linked to the Los Angeles Police Department (LAPD)?
- How much has the city paid out in LAPD-related legal cases since 2005?
- How much was paid out to two men who were freed after decades in prison, along with lawsuits over arrests, injuries, car crashes and alleged retaliation and discrimination against its own employees? 

```{r}
#4.1 - 4.2

lapd_payouts <- payouts_data %>%
  filter(department == 'Police Department') %>%
  summarize(total_amount = sum(amount, na.rm = TRUE))

print(lapd_payouts)

#This cannot be true.. the outcome is 183.252
lapd_percentage <- (lapd_payouts / legal_settlement_payments) * 100


print(lapd_percentage)

#4.3


wrongful_imprisonment <- payouts_data %>%
  filter(case_type == "Police - False Arrest/Detention/Imprisonment") %>%
  summarize(total_amount = sum(amount, na.rm = TRUE)) 

print(wrongful_imprisonment)
  
  

#for employee discrimination - 
various_lapd_lawsuits <- payouts_data %>%
  filter(department == "Police Department"
           case_type == "Law Enforcement Related Injury", "Traffic Accident", , "Retaliation") %>%
 summarize(total_amount = sum(amount, na.rm = TRUE))

print(various_lapd_lawsuits)


  
  
  
```


"Many cases have arisen from people being wounded or killed while crossing city streets: In recent years, the city agreed to pay $15 million to end a legal battle over a man who was gravely injured while crossing Franklin Avenue in Hollywood, another $15 million to the family of two girls hit in a North Hollywood crosswalk, and $9.5 million to the family of a teenage girl fatally struck by a taxi near Dockweiler State Beach." 

#Q5: 

- How much has the city paid out in legal settlements related to pedestrian injuries or fatalities?
- What is the total value of settlements involving people hit while crossing city streets?
- How many legal cases in the dataset involve pedestrian accidents, and what is their combined cost?
- What are the largest individual payouts related to pedestrian incidents in the dataset?

```{r}

#5.1

pedestrian_injuries_payment <- payouts_data %>%
  filter(case_type == 'Traffic Accident - Pedestrian' | case_type == 'Law Enforcement Related Injury') %>%
   summarize(total_amount = sum(amount, na.rm = TRUE))

print(pedestrian_injuries_payment)
  

#I don't have any geo locations in data
geo_cases <- payouts_data %>%
  filter(str_detect(case_name, "Franklin Avenue|Hollywood|Dockweiler|North Hollywood")) %>%
  select(case_name, amount, case_type, department)

print(geo_cases)


```

