---
title: "reverse_engineering_02_Teo"
author: "Teona"
date: "2025-03-10"
output: html_document
---

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(openxlsx)
library(readxl)

```


```{r}
payouts_data <- clean_names(read_excel("input/payouts.xlsx"))

print(payouts_data)
```


"The Times analyzed a city database listing more than 18,000 payouts dating back to July 2005, obtained through a public records request." The problem is that there are no official numbers from before 2005, so reporters can’t really check if what officials are saying is true or not.


#Q1: What are the total legal payout costs in L.A. for each year, and which year had the highest payout?
```{r}
expenses_by_year <- payouts_data %>%
  group_by(fiscal_year) %>%
  summarize(amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(amount))


print(expenses_by_year)
```

Article is published in 2018. 

"Last budget year (2017), the city paid out more than $200 million in legal settlements and court judgments — a record amount that was more than the city spent on its libraries or fixing its streets."

#Q2: 

- How much did the city spend on legal settlements and court judgments in 2017?
- What was the largest amount the city has ever paid out in legal settlements and judgments in a single year?

```{r}
#2.1

legal_settlement_payments <- payouts_data %>%
  filter(fiscal_year == 2017) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE))

print(legal_settlement_payments)

#2.2

largest_settlement_payment <- expenses_by_year %>%
   arrange(desc(amount))


print(largest_settlement_payment)

```

"Last budget year, a class-action lawsuit over a telephone tax cost the city $45 million."

#Q3: 

How much did the class-action lawsuit over a telephone tax cost the city in the last budget year?

```{r}

tele_tax <- payouts_data %>%
  filter(case_type == "Utility Tax") %>%
  summarize(total_amount = sum(amount, na.rm = TRUE))

print(tele_tax)
  

#As I was searching "tax" in payouts_data, only five results popped up and was easy to pick the right one for the filter. Amount paid and the case_type matched, so, the person who received more than $45 milliion for a telephone tax lawsuit was Ardon Estuarado in 2017. 
```

"Thousands of legal battles involving the Los Angeles Police Department, including a rash of recent settlements over police shootings and wrongful imprisonment, have added up to more than 40% of the total — a higher sum than any other city department. That includes more than $24 million paid to two men who were freed after decades in prison, along with lawsuits over arrests, injuries, car crashes and alleged retaliation and discrimination against its own employees."

#Q4: 

- What percentage of the city’s total legal payouts are linked to the Los Angeles Police Department (LAPD)?
- How much has the city paid out in LAPD-related legal cases since 2005?
- How much was paid out to two men who were freed after decades in prison, along with lawsuits over arrests, injuries, car crashes and alleged retaliation and discrimination against its own employees? 

```{r}
#4.1 - 4.2

lapd_payouts <- payouts_data %>%
  filter(department == 'Police Department') %>%
  summarize(total_amount = sum(amount, na.rm = TRUE))

print(lapd_payouts)


total_legal_settlements <- payouts_data %>%
   summarize(total_amount = sum(amount, na.rm = TRUE))


lapd_percentage <- (lapd_payouts$total_amount / total_legal_settlements$total_amount) * 100


print(lapd_percentage)

#4.3

wrongful_imprisonment <- payouts_data %>%
  filter(case_type == "Police - False Arrest/Detention/Imprisonment") %>%
  summarize(total_amount = sum(amount, na.rm = TRUE)) 

print(wrongful_imprisonment)
  
  

arrests_injuries_crashes <- payouts_data %>%
  filter(filter(case_type ==
    "Police - Civil Rights Violation",
    "Traffic Accident - Pedestrian",
    "Traffic Accident - Other",
    "Traffic Accident - Failed to Stop/Yield"
  )) %>%
  summarize(arrests_injuries_crashes = sum(amount, na.rm = TRUE))
  
  print(arrests_injuries_crashes)


retaliation_discrimination <- payouts_data %>%
  filter(case_type ==
    "Retaliation",
    "Gender Discrimination",
    "Race/Color/National Origin Discrimination") %>%
  summarize(total_retaliation_discrimination = sum(amount, na.rm = TRUE))


print(retaliation_discrimination)
```


Most common payouts in L.A. 

#Q5: 

- What is the percentage breakdown of legal payouts (2005-2018) by category?

```{r}
payout_document <- clean_names(read_csv("output/type-totals.csv"))
```




```{r}

top_payouts <- payout_document %>%
  arrange(desc(amount_percent)) %>%
  slice(1:6)


ggplot(top_payouts, aes(x = reorder(latimes_type, amount_percent), y = amount_percent)) +
  geom_col(fill = "#1f77b4") +
  coord_flip() +
  labs(
    title = "Top 6 Payout Categories by Percentage",
    x = "Category",
    y = "Percent of Total Payouts"
  ) +
   theme_economist()

```
