---
title: "lab_07"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}

census_api_key("090efdf684b528f0a4a9fc2fbaafe6e556f4ab18", install=TRUE)

```

## Load and view ACS variables

```{r}

sf1 <- load_variables(2010, "sf1", cache = TRUE)
View(sf1)

```


You'll also need to load the list of American Community Survey variables from the 2023 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
acs5 <- load_variables(2021, "acs5", cache = TRUE)

```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2023 5-year estimates? You will need to find the variable for median age first. **A1**

```{r}
median_age <- get_acs(geography = "county",
              variables = c(medage = "B01002_001"),
              state = "MD",
              year = 2023)

```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. **A2**

```{r}



get_acs(geography = "county",
              variables = c(medage = "B01002A_001"),
              state = "MD",
              year = 2023)


```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1.  Load the Montgomery County 911 overdose calls data.
2.  Create a new dataframe with the total number of calls for each zip code.
3.  Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).
4.  Get from the ACS data from all Maryland zip codes using that variable - you will need to specify an earlier year to get the data.
5.  Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
6.  Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

Which zip code has the highest rate of calls? Which has the lowest?

**A3**

```{r}

montgomery_calls <- read_csv("data/montgomery_2022_overdoses.csv") %>%
  mutate(zip = as.character(zip)) %>%
  group_by(zip) %>%
  summarise(total_calls = n())
  

# Get data from the ACS for population 18 and up by zip code in MD. 2019 is the latest year that has data

pop18_and_up_by_zip <- get_acs(
  geography = "zcta", 
          variables = c(pop18_and_up = "DP05_0021E"), 
          state="MD", 
          year=2019
          )

# Join the Montgomery overdose calls with the ACS population 18 and up data, rename columns and
# calculate overdose calls per 100,000 people 18 and up

calls_by_zip_with_pop18_and_up <- montgomery_calls %>%
  left_join(pop18_and_up_by_zip, by=c('zip'='GEOID')) %>%
  rename(pop18_and_up = estimate) %>%
  mutate(call_rate = total_calls / pop18_and_up * 100000)

# Arrange by most calls per 100,000
calls_by_zip_with_pop18_and_up %>%
  arrange(desc(call_rate))

# Arrange by fewest calls per 100,000
calls_by_zip_with_pop18_and_up %>%
  arrange(call_rate)
```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4**

#ZIP code 20877 has a total population of 39,068, with a median age of 37.3. The per capita income is $38,859, while the median household income is $86,103. About 11.3% of the population lives in poverty. There are 12,880 households, with an average of 3 persons per household, and 47% of residents are married.

#ZIP code 20707 has a total population of 36,549, with a median age of 39.1. The per capita income is $49,777, while the median household income is $104,093. Approximately 7.5% of the population lives in poverty. There are 14,321 households, with an average of 2.5 persons per household, and 46% of residents are married.


#There appears to be a correlation between household income and overdose calls, with higher-income neighborhoods reporting fewer incidents. It would be interesting to compare factors like education and lifestyle between these areas and explore how the environment influences life choices, including drug use.

```{r}
highest_rates <- calls_by_zip_with_pop18_and_up %>%
  arrange(desc(call_rate)) 

print(highest_rates)


lowest_rates <- calls_by_zip_with_pop18_and_up %>%
  filter(total_calls > 0) 

print(lowest_rates)
 
```

**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

**A5**

#No, I got two different zip codes than before. I guess, zio codes here should have been the same as for Q3. 
```{r}
calls_by_zip <- montgomery_calls %>%
  left_join(pop18_and_up_by_zip, by = c('zip' = 'GEOID')) %>%
  rename(pop18_and_up = estimate) %>%
  
  mutate(pop_min = pop18_and_up - moe,
    pop_max = pop18_and_up + moe)
  
  
  calls_by_zip %>%
  arrange(desc(total_calls)) %>%
  select(pop_min, pop_max, zip) 
    

  
  
  
  
#I’m going to join the data together, so each county has one line of data. Then I’m going to rename some fields that repeat. Then I’m going to calculate the minimium and maximum value of the estimate using the margin of error. That’ll help me later. After that, I’m going to calculate a perent change and sort it by that change.
  
#rename(estimate2010=estimate.x, estimate2020=estimate.y) |>
  mutate(min2010 = estimate2010-moe.x, max2010 = estimate2010+moe.x, min2020 = estimate2020-moe.y, max2020 = estimate2020+moe.y) |>
  select(-variable.x, -variable.y, -moe.x, -moe.y) |>
  mutate(change = ((estimate2020-estimate2010)/estimate2010)*100) |>
  arrange(change)

```
